---
title: "Exercise set: Sampling distributions"
author: "BIOL8001 Graduate Statistics"
date: "2026-09-16"
output:
  github_document:
    html_preview: True
  pdf_document:
    latex_engine: xelatex
knit: (function(input, ...) rmarkdown::render(input, output_format = "all"))
mainfont: "Gentium Book Plus"
geometry: margin=2.5cm
urlcolor: blue
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r penguins}
population <- c(12, 34, 69, 36, 13, 47, 98, 12, 18, 87, 65,
                42, 98, 76, 34, 18, 10, 89, 90, 78, 65, 53,
                27, 57, 69, 23, 44, 12, 67, 11, 15)
```

For the calculations in this sheet you can use R, the scripts from class, or two online calculators: one for the [normal distribution](https://mabognar.github.io/apps/normal.html) and one for the [t distribution](https://mabognar.github.io/apps/t.html).

## Question 1: The width of the sampling distribution

As we have seen, the standard deviation of the sampling distribution of the mean is

$$\sigma_{\bar{x}} = \frac{\sigma}{\sqrt{n}}$$

which is smaller than the spread of the population itself for any sample size bigger than one. The left panel shows a colony with μ = 80 kg and σ = 12 kg. The right panel plots the sd of the sampling distribution against `n` for that colony.

```{r q1-se, fig.width=7, fig.height=3.2}
sigma <- 12
mu <- 80
par(mfrow = c(1, 2), mar = c(4.2, 4, 2.6, 1), cex.main = 1, font.main = 1)

curve(dnorm(x, mu, sigma), from = mu - 4 * sigma, to = mu + 4 * sigma, n = 400,
      lwd = 3, col = "grey30", xlab = "Weight (kg)", ylab = "Density",
      main = expression(paste("A Colony: ", mu, " = 80 kg, ", sigma, " = 12 kg")))
abline(v = mu, lty = 2)
arrows(mu, dnorm(mu + sigma, mu, sigma), mu + sigma, dnorm(mu + sigma, mu, sigma),
       length = 0.08, col = "red", lwd = 2)
text(mu + sigma / 2, dnorm(mu + sigma, mu, sigma), expression(sigma), pos = 3, col = "red")

nn <- 1:60
plot(nn, sigma / sqrt(nn), type = "l", lwd = 3, col = "grey30", ylim = c(0, 13),
     xlab = "Sample Size n", ylab = "SD of Sample Means (kg)",
     main = "Width of the Sampling Distribution")
points(c(2, 5, 10, 20, 40), sigma / sqrt(c(2, 5, 10, 20, 40)), pch = 19, col = "red")
abline(h = sigma, lty = 3)
text(60, sigma, expression(paste("population: ", sigma, " = 12 kg")), adj = c(1, -0.4), cex = 0.8)
```

**(a)** Take the colony in the left panel. A single penguin weighing 90 kg is nothing unusual there, since it is less than one σ above the mean. Now think about what has to happen for a sample of five penguins from this colony to average 90 kg. Use the comparison to explain why the sampling distribution is narrower than the population. Do not quote the formula back.

**(b)** The curve drops steeply and then flattens. Using the figure or the formula: how many penguins do you need to halve the width you get at `n = 10`? And to halve it again?

**(c)** Each row below is a different variable in a different population. The first four columns describe the population. The last three describe the sampling distribution of the mean for samples of size `n`.

| Variable             | population mean μ | population sd σ | population shape | sample size n | mean of the sampling distribution | sd of the sampling distribution | shape of the sampling distribution |
| -------------------- | ----------- | ----------- | ----------- | -------- | ------------- | ------------- | ------------- |
| Flipper length (mm)  | 190         | 8           | normal      | 16       |               |               |               |
| Egg mass (g)         | 95          | 12          | normal      | 9        |               |               |               |
| Dive depth (m)       | 40          | 15          | strongly skewed | 25       |               |               |               |
| Birds per colony     | 500         | 100         | strongly skewed | 4        |               |               |               |

Fill in the last three cells. Mark each entry as exact, approximate, or unknown, and give your reason for any entry that is not exact.

**(d)** For row 4, you want the probability that a sample of four colonies averages more than 600 birds. Can you get it from a normal curve (R or the normal calculator) with the mean and sd in your table? Could you answer the same kind of question for row 3 that way, say the probability that the mean dive depth exceeds 45 m?

## Question 2: Four colonies that agree on μ and on σ

The top row shows four colonies of 2000 penguins each. All four have a mean of exactly 50 kg and a standard deviation of exactly 12 kg. The second and third rows show the sampling distribution of the mean for each colony at two sample sizes. The figure is made the same way as `SameMeanSameSD.R`.

```{r q2-shapes, fig.width=7, fig.height=6.6}
set.seed(4003)
rescale <- function(x, m, s) (x - mean(x)) / sd(x) * s + m
N <- 2000
colonies <- list(
  "Colony A" = rescale(rnorm(N), 50, 12),
  "Colony B" = rescale(runif(N), 50, 12),
  "Colony C" = rescale(c(rnorm(N / 2, -1, 0.35), rnorm(N / 2, 1, 0.35)), 50, 12),
  "Colony D" = rescale(rbeta(N, 1.5, 6), 50, 12))

draw_means <- function(x, n, reps = 20000) replicate(reps, mean(sample(x, n, replace = TRUE)))

par(mfrow = c(3, 4), mar = c(4, 3.4, 2.4, 0.8), cex.main = 1)

pop_breaks <- seq(-10, 110, by = 2)
mean_breaks <- seq(0, 100, by = 0.5)

for (nm in names(colonies))
{
  hist(colonies[[nm]], breaks = pop_breaks, xlim = c(10, 90), col = "grey45",
       border = NA, main = "", xlab = "kg", ylab = "")
  title(main = nm, line = 1.4)
  mtext(bquote(mu == .(round(mean(colonies[[nm]]))) ~ "kg," ~~
               sigma == .(round(sd(colonies[[nm]]))) ~ "kg"),
        side = 3, line = 0.2, cex = 0.7)
}

for (nm in names(colonies))
  hist(draw_means(colonies[[nm]], 2), breaks = mean_breaks, xlim = c(20, 80),
       col = "grey45", border = NA,
       main = paste(nm, ": mean of 2"), xlab = "kg", ylab = "")

for (nm in names(colonies))
  hist(draw_means(colonies[[nm]], 30), breaks = mean_breaks, xlim = c(20, 80),
       col = "grey45", border = NA,
       main = paste(nm, ": mean of 30"), xlab = "kg", ylab = "")
```

**(a)** The four colonies agree on both numbers we normally use to summarise a population. What has the summary thrown away? Using the top row, describe for one colony of your choice something a biologist would want to know that μ and σ do not tell them. How does this relate to Stephen J Gould’s essay [The Median Isn’t the Message](https://journalofethics.ama-assn.org/article/median-isnt-message/2013-01)?

**(b)** Colony C has a mean of 50 kg, and almost no penguin in it weighs anything close to 50 kg. Its sampling distribution of the mean is nevertheless centred on 50. Is the sampling distribution then centred on a lie? Say what the number 50 is a correct statement about, and what it is not.

**(c)** Here are two separate statements about the sampling distribution of the mean: (1) its mean is μ and its standard deviation is σ/√n, as in Question 1, and (2) its shape is approximately normal, which is what the central limit theorem says for large samples. Compare the four panels in the middle row with each other, then compare the middle row with the bottom row. For each claim, say whether it holds for all four colonies at both sample sizes. One of the two claims does not always hold. Name the panel where it fails most clearly, and explain why that panel looks the way it does.

**(d)** A colleague says: "My measurements are clearly not normally distributed, so none of this applies to my data." Which distribution does the normality claim in (c) refer to: the population, the one sample they collected, or the sampling distribution? Answer using the figure, and say what you would tell the colleague.

## Question 3: The mean is not the only statistic

A statistic is any number you calculate from a sample, so the mean is not special. The median, the sd, the minimum and the maximum all have sampling distributions too; `PenguinStatistics.R` draws them for the penguins in `PenguinMean.R`. This question looks at one of them more closely: the maximum. Each of the two figures below shows three colonies of 2000 penguins (top row), the sampling distribution of the heaviest penguin in a sample of five (middle row), and the cumulative distributions of that heaviest penguin for all three colonies together (bottom panel). A cumulative distribution gives, for every weight on the horizontal axis, the proportion of samples whose heaviest penguin is at or below that weight. The dashed black lines mark the heaviest penguin in each colony. The red line marks a researcher's observation: five penguins, the heaviest of which weighs 70 kg.

```{r q3-max-setup}
set.seed(4030)
colony_size <- 2000
catch <- 5
max_reps <- 20000
observed_max <- 70
max_cols <- c("steelblue", "darkorange", "forestgreen")

# Colonies are built from evenly spaced quantiles, so their shapes are exact
# rather than one random draw.
stretch <- function(x, low, high) low + (high - low) * (x - min(x)) / (max(x) - min(x))
colony <- function(a, b, low, high) stretch(qbeta(ppoints(colony_size), a, b), low, high)
draw_max <- function(x) replicate(max_reps, max(sample(x, catch, replace = TRUE)))

max_figure <- function(colonies, maxima, heading)
{
  layout(matrix(c(1, 2, 3, 4, 5, 6, 7, 7, 7), nrow = 3, byrow = TRUE),
         heights = c(1, 1, 1.3))
  par(mar = c(4, 3.6, 2.8, 0.8), cex.main = 1, font.main = 1,
      oma = c(0, 0, 2, 0), mgp = c(2.4, 0.8, 0))
  breaks <- seq(20, 100, by = 2)

  for (i in seq_along(colonies))
  {
    hist(colonies[[i]], breaks = breaks, xlim = c(0, 110), col = max_cols[i],
         border = NA, main = names(colonies)[i], xlab = "Weight (kg)", ylab = "")
    abline(v = max(colonies[[i]]), lty = 2)
  }

  for (i in seq_along(colonies))
  {
    hist(maxima[[i]], breaks = breaks, xlim = c(0, 110), col = max_cols[i],
         border = NA, main = "Heaviest of 5", xlab = "Weight (kg)", ylab = "")
    abline(v = max(colonies[[i]]), lty = 2)
  }

  plot(NA, xlim = c(0, 110), ylim = c(0, 1), xlab = "Heaviest of 5 (kg)",
       ylab = "Proportion at or Below",
       main = "Cumulative Distributions of the Heaviest of 5")
  for (i in seq_along(colonies))
    lines(ecdf(maxima[[i]]), col = max_cols[i], lwd = 2.5,
          do.points = FALSE, verticals = TRUE)
  abline(v = observed_max, lty = 2, col = "red", lwd = 2)
  legend("topleft", bty = "n", lwd = 2.5, col = max_cols, legend = names(colonies))
  mtext(heading, outer = TRUE, cex = 1.1)
}

# Chance that all five penguins weigh observed_max or less.
p_at_or_below <- function(colonies)
  sprintf("%.2f", sapply(colonies, function(x) mean(x <= observed_max)^catch))

colonies_1 <- list("Heaviest 60 kg" = colony(1, 1, 20, 60),
                   "Heaviest 80 kg" = colony(1, 1, 20, 80),
                   "Heaviest 100 kg" = colony(1, 1, 20, 100))
maxima_1 <- lapply(colonies_1, draw_max)
p_1 <- p_at_or_below(colonies_1)

colonies_2 <- list("Mostly Light" = colony(1, 4, 20, 100),
                   "Even" = colony(1, 1, 20, 100),
                   "Mostly Heavy" = colony(4, 1, 20, 100))
maxima_2 <- lapply(colonies_2, draw_max)
p_2 <- p_at_or_below(colonies_2)
```

In the first figure, all three colonies have the same shape: weights are spread evenly from 20 kg up to the heaviest penguin, which weighs 60, 80 or 100 kg.

```{r q3-max-maxima, fig.width=7.5, fig.height=7.5}
max_figure(colonies_1, maxima_1, "Same Shape, Different Heaviest Penguin")
```

**(a)** Compare the top and middle rows. Where does the heaviest of five sit relative to the heaviest penguin in the colony? Can a sample of five ever produce a value on the other side of the dashed line? Suppose you had to report the heaviest penguin in a colony and could only catch five. What would you tell a reader about your number? Would catching 20 remove the problem, or only shrink it?

**(b)** Use the bottom panel to answer these for the researcher whose heaviest penguin weighs 70 kg:

  1. Which colony can you rule out completely, and why?
  2. For the colony whose heaviest penguin weighs 80 kg, read off the proportion of samples whose heaviest penguin is 70 kg or lighter. Do the same for the 100 kg colony. (The values are `r p_1[2]` and `r p_1[3]`.)
  3. Write one sentence that says exactly what `r p_1[3]` is the probability of. Make clear what is assumed and what is observed. Compare your sentence with the one you write in Question 6(a).

In the second figure, all three colonies range from exactly 20 kg to exactly 100 kg, so they share the same heaviest penguin. They differ only in how the weights in between are spread.

```{r q3-max-shapes, fig.width=7.5, fig.height=7.5}
max_figure(colonies_2, maxima_2, "Same Heaviest Penguin (100 kg), Different Shapes")
```

**(c)** For each of these three colonies, read off the proportion of samples whose heaviest penguin is 70 kg or lighter. (The values are `r p_2[1]`, `r p_2[2]` and `r p_2[3]`.) All three colonies have a heaviest penguin of 100 kg. Why are the answers so different? A colleague says: "Our heaviest penguin was only 70 kg, so the heaviest penguin in the colony is probably not 100 kg." What would the colleague have to assume about the colony to back this up?

**(d)** Only the mean has a tidy formula (Question 1) for its sampling distribution. None of the curves in this question came from a formula. How were they made? (`PenguinStatistics.R` does the same for the mean, median, sd, minimum and maximum. Say in one sentence what it is doing.)

**(e)** (Optional) For the colonies in the first figure, weights are spread evenly between 20 kg and the heaviest penguin M, so the chance that one penguin weighs 70 kg or less is (70 − 20)/(M − 20). Use this to work out the chance that all five penguins in a sample weigh 70 kg or less, and check your answer against the values in (b).

## Question 4: When σ is unknown

Everything so far assumed we knew the population standard deviation σ. We almost never do. The first figure shows the t distribution: the sampling distribution of

$$t = \frac{\bar{x} - \mu}{s/\sqrt{n}}$$

for samples of five penguins (`df` = 4), with a standard normal drawn on top of it for comparison. Both are on the t axis, which counts standard errors rather than kilograms. See `TCurve.R`.

```{r q4-t, fig.width=6.5, fig.height=3.1}
par(mar = c(4.2, 4, 2.8, 1))
curve(dt(x, df = 4), from = -5, to = 5, n = 500, lwd = 3, col = "grey30",
      xlab = "t  (standard errors from the centre)", ylab = "density",
      main = "t with df = 4, and a normal")
curve(dnorm(x), add = TRUE, lwd = 2, lty = 2, col = "red")
legend("topright", c("t, df = 4", "normal"), lwd = c(3, 2), lty = c(1, 2),
       col = c("grey30", "red"), bty = "n", cex = 0.85)
```

The second figure shows the same t-curve with a kilogram ruler, for a sample of five penguins with an observed sd of 12 kg and an assumed mean of 50 kg. See `TAxisInKilograms.R`.

```{r q4-rulers, fig.width=6.5, fig.height=3.3}
mu <- 50; s <- 12; n5 <- 5
se <- s / sqrt(n5)
par(mar = c(4.2, 4, 5, 1))
curve(dt((x - mu) / se, df = 4) / se, from = mu - 5 * se, to = mu + 5 * se, n = 500,
      lwd = 3, col = "grey30", xlab = "kg", ylab = "density", xaxt = "n", main = "")
title(main = "the same curve, two rulers", line = 3.4)
axis(1, at = mu + (-3:3) * se, labels = round(mu + (-3:3) * se, 1))
axis(3, at = mu + (-3:3) * se, labels = -3:3)
mtext("t  (standard errors)", side = 3, line = 1.8, cex = 0.9)
```

**(a)** In your own words, what problem does the t distribution solve? Your answer should name the piece of information we lost and say what we put in its place.

**(b)** Compare the two curves in the first figure. Where do they agree, and where do they differ? For a normal population, $\bar{x}$ has an exactly normal sampling distribution. Why doesn't t? The t statistic uses the sample sd instead of σ, and the sample sd changes from sample to sample. Use that to explain why the t curve differs from the normal in the way it does. (Hint: what happens to t when a sample happens to give a sample sd much smaller than σ?)

**(c)** The hint in (b) points at samples whose sd comes out smaller than σ. How often does that happen?

  1. Before calculating anything, guess: if you draw many samples of five from a normal population with σ = 12 kg, what fraction of them will have a sample sd below 12 kg?
  2. Check your guess by simulation. Draw 10,000 samples of five with `rnorm(5, 50, 12)`, calculate the `sd()` of each, and count how often it is below 12. Also calculate the average of the 10,000 sds, and the average of their squares (the variances).
  3. Compare the average sd with σ = 12, and the average variance with σ² = 144. Explain how both results can be true at once. (Hint: how far below σ can a sample sd go, and how far above it?)
  4. Repeat with samples of 2 and of 30. How does the fraction below σ change, and what does that suggest about which samples fill the tails of the t curve when `n` is small?

**(d)** For a normal distribution, 95% of the curve lies within about 1.96 standard errors of the centre. Find the corresponding number for the t with `df` = 4 (use `TCurve.R`, R, the t calculator, or a table). Then use the second figure to say what the difference between those two numbers is worth **in kilograms** for this sample. Would you have called that difference negligible before computing it?

**(e)** Run `TCurve.R` with `n <- 5` and then with `n <- 30`, and compare the two curves. What happens, and why does the distinction between the t and the normal stop mattering as the sample grows? Connect this to what the sample sd is doing as `n` increases.

## Question 5: From a sampling distribution to a confidence interval

We did not cover this in class, but the questions contain sufficient information to solve them.

A researcher catches five penguins from a large colony. Weights in this colony are normally distributed, but the researcher knows neither the colony's mean nor its sd. The five weights are 89, 78, 65, 44, and 34 kg, so the sample mean is 62 kg, and the sample sd is about 22.9 kg. The researcher writes: "The mean weight of this colony is 62 kg."

**(a)** What is wrong with that sentence? Think about what a second researcher would report after catching five different penguins. Of the three kinds of distribution we have met (the population, the one sample, the sampling distribution), which one tells you how far a sample mean typically lands from the colony's true mean? The researcher only has one sample. What can they calculate from it to estimate the width of that distribution?

**(b)** Recall the t statistic from Question 4:

$$t = \frac{\bar{x} - \mu}{s/\sqrt{n}}$$

Question 4(d) gave the value $t_{0.975,\,n-1}$ that leaves 95% of the `df` = 4 curve in the middle. So in 95% of samples of five, t lands between $-t_{0.975,\,n-1}$ and $t_{0.975,\,n-1}$. Multiply through by $s/\sqrt{n}$, and this says that the sample mean lands within $t_{0.975,\,n-1}$ estimated standard errors of μ:

$$|\bar{x} - \mu| \le t_{0.975,\,n-1} \cdot \frac{s}{\sqrt{n}}$$

The standard error $s/\sqrt{n}$ is calculated from that same sample. Now turn this around. If $\bar{x}$ is within that distance of μ, then μ is within the same distance of $\bar{x}$. This lets us build an interval around the one number we actually have:

$$\bar{x} \pm t_{0.975,\,n-1} \cdot \frac{s}{\sqrt{n}}$$

In the sentence "in 95% of samples, μ lies within this distance of $\bar{x}$", which quantities change from sample to sample, and which stays fixed?

**(c)** Work out this 95% confidence interval for the researcher's five penguins, in three steps:

  1. The margin of error, $t_{0.975,\,n-1} \cdot s/\sqrt{n}$.
  2. The lower limit of the interval.
  3. The upper limit of the interval.

Then rewrite the researcher's sentence so that it is defensible.

**(d)** The figure below repeats the researcher's study 50 times. Each time it draws five penguins from the same colony and builds the interval from (c). For the simulation, the colony has a mean of 50 kg and an sd of 20 kg. The dashed line marks that true mean of 50 kg, which a real researcher would never get to see. Intervals that miss it are drawn in red.

```{r q5-ci, fig.width=6.5, fig.height=5}
set.seed(4005)
n <- 5
reps <- 50
true_mean <- 50
true_sd <- 20
t_crit <- qt(0.975, df = n - 1)
cis <- t(replicate(reps, {
  x <- rnorm(n, true_mean, true_sd)
  mean(x) + c(-1, 0, 1) * t_crit * sd(x) / sqrt(n)
}))
miss <- cis[, 1] > true_mean | cis[, 3] < true_mean

par(mar = c(4.2, 4, 2.6, 1))
plot(NA, xlim = range(cis), ylim = c(1, reps), yaxt = "n",
     xlab = "kg", ylab = "repeat of the study", main = "50 Samples of Five, 50 Intervals")
axis(2, at = c(1, 10, 20, 30, 40, 50), las = 1)
segments(cis[, 1], 1:reps, cis[, 3], 1:reps, lwd = 2,
         col = ifelse(miss, "red", "grey40"))
points(cis[, 2], 1:reps, pch = 19, cex = 0.6, col = ifelse(miss, "red", "grey20"))
abline(v = true_mean, lty = 2, lwd = 2)
```

`r sum(miss)` of the 50 intervals miss the true mean. The true mean stays put in every repeat. What changes from one repeat to the next? So what exactly is the "95%" a property of? Is it a property of any single interval in the figure?

**(e)** Say whether each of these readings of the researcher's interval is defensible, and explain what is wrong with the ones that are not. Compare them with the p-values in Question 6(b).

  1. There is a 95% probability that the colony's mean lies inside this interval.
  2. 95% of the penguins in the colony weigh something inside this interval.
  3. If many researchers each caught five penguins and built an interval this way, about 95% of their intervals would contain the colony's mean.
  4. If the researcher caught five new penguins, there is a 95% chance their mean would fall inside this interval.

**(f)** The researcher goes back and catches 20 penguins, and the sample sd comes out about the same. Roughly how much narrower will the interval be? Two things make it narrower. Name both, and use Question 1(b) to say which of them does most of the work. What does this mean for someone deciding how many penguins to catch?

**(g)** The recipe in (b) gives an interval that contains μ in exactly 95% of samples, but only if the colony is normal: Student's theorem, which gives the t distribution, needs that assumption. What happens if the colony is not normal? The figure below takes the four colonies from Question 2, which all have a mean of 50 kg. For each colony and each sample size, it builds 10,000 intervals the way you did in (c), and records the percentage of intervals that contain 50 kg.

```{r q5-coverage, fig.width=6.5, fig.height=3.8}
set.seed(4006)
sizes <- c(2, 3, 5, 10, 20, 30)
coverage <- sapply(colonies, function(x)
  sapply(sizes, function(k)
  {
    tk <- qt(0.975, df = k - 1)
    100 * mean(replicate(10000, {
      s <- sample(x, k, replace = TRUE)
      abs(mean(s) - 50) <= tk * sd(s) / sqrt(k)
    }))
  }))

cov_cols <- c("grey30", "steelblue", "darkorange", "forestgreen")
par(mar = c(4.2, 4.2, 2.6, 1))
matplot(sizes, coverage, type = "b", pch = 19, lty = 1, lwd = 2, col = cov_cols,
        log = "x", xaxt = "n", ylim = c(80, 97),
        xlab = "Sample Size n", ylab = "Intervals Containing 50 kg (%)",
        main = "How Often the 95% Interval Captures the Mean")
axis(1, at = sizes)
abline(h = 95, lty = 2)
legend("bottomright", names(colonies), col = cov_cols, lwd = 2, pch = 19, bty = "n")
```

  1. For which colony is the label "95% interval" exactly right? For the other three, what does the formula give you instead?
  2. Which colony gives the worst intervals, and at which sample sizes? Look back at its sampling distribution for samples of two in Question 2, and suggest why.
  3. Which colony stays below 95% the longest as `n` grows? What feature of its shape might be responsible?
  4. Why do all four colonies get close to 95% for larger samples? Which result from class is doing the work there?
  5. A researcher does not know the shape of their colony and catches five penguins. Can they build an interval that is guaranteed to contain μ in 95% of samples? What are they assuming if they use the formula from (b) anyway?

## Question 6: Using a sampling distribution for inference

This is the calculation from the end of the slides. We assume a population of penguins with μ = 5 kg and σ = 3 kg. We catch 15 of them, and the sample mean is 6.15 kg. The figure shows the sampling distribution under this assumption, with the observed mean marked.

```{r q6-inference, fig.width=6.5, fig.height=3.2}
mu0 <- 5; sig <- 3; n6 <- 15; obs <- 6.15
se6 <- sig / sqrt(n6)
xs <- seq(mu0 - 4 * se6, mu0 + 4 * se6, length.out = 500)
plot(xs, dnorm(xs, mu0, se6), type = "l", lwd = 3, col = "grey30",
     xlab = "mean weight of a sample of 15 penguins (kg)", ylab = "density",
     main = expression(paste("Sampling distribution if  ", mu, " = 5 and ", sigma, " = 3,  n = 15")))
xt <- seq(obs, mu0 + 4 * se6, length.out = 200)
polygon(c(obs, xt, max(xt)), c(0, dnorm(xt, mu0, se6), 0), col = "grey80", border = NA)
lines(xs, dnorm(xs, mu0, se6), lwd = 3, col = "grey30")
abline(v = obs, lty = 2, lwd = 2, col = "red")
text(obs + 0.05, dnorm(mu0, mu0, se6) * 0.6, "observed 6.15", col = "red", adj = 0)
```

The shaded area is `r round(pnorm(obs, mu0, se6, lower.tail = FALSE), 5)`.

**(a)** Write one sentence that says exactly what that number is the probability of. Your sentence must make clear what is being assumed and what is being observed, and it should be possible to tell from it that the number is not a statement about a single penguin.

**(b)** Here are four readings of the same number. Decide for each whether it is defensible, and say what is wrong with the ones that are not.

  1. There is a 6.9% chance that the population mean is 5 kg.
  2. There is a 6.9% chance that the difference between 6.15 kg and 5 kg is just sampling variation.
  3. If μ really were 5 kg and σ really were 3 kg, and many research groups each weighed 15 penguins, about 7% of them would report a mean of 6.15 kg or more.
  4. The probability of getting a sample mean of exactly 6.15 kg is 6.9%.

**(c)** The question asked was about a mean of "6.15 kg **or larger**". Why is it phrased that way rather than asking about 6.15 kg itself? (Question 4 of the previous exercise set is the relevant one here.)

**(d)** Keeping the same 15 penguins and the same observed mean of 6.15 kg, the slides redo the calculation assuming μ = 3 instead, and get `r format(round(pnorm(obs, 3, se6, lower.tail = FALSE), 5), scientific = FALSE)`. Nothing about the penguins changed between the two calculations. So what is the p-value a property of? Name every ingredient it depends on.

**(e)** Suppose the same mean of 6.15 kg had come from 60 penguins instead of 15, with μ = 5 and σ = 3 as before. Work out the new probability (`NormalCurve.R` will draw it, or use R or the normal calculator). What changed, and what did not? What does this suggest about reporting a p-value without the sample size next to it?

**(f)** The slides stop here with the words "now what?" and take no decision. To turn `r round(pnorm(obs, mu0, se6, lower.tail = FALSE), 5)` into a decision about the assumed population (keep μ = 5 kg or reject it), you need a cut-off. What is that cut-off usually called, and what does it stand for? Is it anywhere in the data, and if not, where does it come from?

## Question 7: A statistic built from two samples

This question goes beyond what we did in class, and it is the one to attempt last.

A questionnaire asks whether animal research is wrong, answered on a 7-point scale. Assume that in the population the mean for women is 5, the mean for men is 4, and the standard deviation in both groups is 1.55. Twelve women and twelve men are selected at random, and we record the difference between the two sample means.

**(a)** That difference is a statistic, calculated from a sample, so it has a sampling distribution like everything else in this sheet. Before calculating anything: where would that distribution be centred, and would you expect it to be wider or narrower than the sampling distribution of either group's mean on its own? Say why.

**(b)** First write down the sampling distribution of each group's mean: its mean, its sd and its shape. Then look up how the difference between two independent normal variables is distributed (for example in the Wikipedia article on the [sum of normally distributed random variables](https://en.wikipedia.org/wiki/Sum_of_normally_distributed_random_variables)), and use it to get the sampling distribution of the difference. What is the probability that the women's sample mean comes out more than 1.5 points above the men's? The variances add even though you are subtracting the means. Why does subtracting not make the spread smaller?

**(c)** Check your answer by simulation. The code below draws 12 women and 12 men from normal populations with the parameters above, records the difference in means, and repeats this 10,000 times. Run it. How close is the simulation to your calculation, and what would make it closer?

```{r q7-code, echo=TRUE, eval=FALSE}
repeats <- 10000
differences <- numeric(repeats)
for (i in 1:repeats)
{
  women <- rnorm(12, mean = 5, sd = 1.55)
  men <- rnorm(12, mean = 4, sd = 1.55)
  differences[i] <- mean(women) - mean(men)
}
hist(differences, breaks = 50)
mean(differences > 1.5)
```

**(d)** A study of this size finds a difference of 1.5 points. We know the true difference is 1 point, but the researchers do not. Using the idea from Question 5, build a 95% interval around their 1.5 (σ is known here, so use the normal distribution rather than the t). Which population differences are compatible with their result? What does this say about how much a single study of 12 and 12 tells you about the size of the difference?
