---
title: "Distributions as Shortcuts: Building from Process to Distribution"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5)
```

# Overview

This lab demonstrates the core idea: **a probability distribution is the mathematical shortcut for a process we could have simulated, given its parameters**.

For each example, we: 1. Simulate the process manually from first principles 2. Visualize the resulting distribution 3. Reveal the R distribution function as the shortcut 4. Overlay the shortcut on the manual simulation to show they match

Each example takes \~15 minutes in class: run the manual code, discuss the visualization, then introduce the shortcut.

------------------------------------------------------------------------

# Example 1: Binomial — Counting successes in n trials

**Process**: Flip a biased coin (p=0.6) 20 times. How many heads?

**Teaching moment**: The binomial distribution is the shortcut for counting successes in n independent Bernoulli trials.

```{r binomial-manual, fig.cap="Manual simulation: 20 coin flips, p=0.6"}
n <- 20
p <- 0.6
n_reps <- 10000
set.seed(12345)

# Manual: sample 0/1, n times, count 1s, repeat
results_manual <- numeric(n_reps)
for (i in 1:n_reps) {
  flips <- sample(c(0, 1), n, replace = TRUE, prob = c(1 - p, p))
  results_manual[i] <- sum(flips)
}

hist(results_manual, breaks = seq(0, n) + 0.5, freq = FALSE,
     main = "Manual: 20 coin flips, p = 0.6",
     xlab = "Number of heads")
```

```{r binomial-shortcut, fig.cap="Binomial shortcut (dbinom) overlaid on manual simulation"}
# Shortcut: the binomial formula, dbinom(), evaluated at each possible count
k <- 0:n

hist(results_manual, breaks = seq(0, n) + 0.5, freq = FALSE,
     main = "Manual (black bars) vs Shortcut (red points)",
     xlab = "Number of heads")
points(k, dbinom(k, n, p), col = "red", pch = 19)
legend("topright", legend = c("Manual simulation", "dbinom() shortcut"),
       col = c("black", "red"), pch = c(15, 19), bty = "n")
```

------------------------------------------------------------------------

# Example 2: Poisson — Rare events in fixed interval

**Process**: Customers enter a store at average rate λ=3 per hour. How many in one hour?

**Teaching moment**: The Poisson distribution is the shortcut for counting rare events in a fixed interval.

*Note:* the "manual" version here builds the count as the **limit of a binomial** — chop the hour into many ($n = 10{,}000$) tiny time-slices, each with a tiny probability of an event, and count how many fire. Poisson emerges as that limit; it is not a from-scratch mechanism of its own.

```{r poisson-manual, fig.cap="Binomial-limit simulation: rare events (lambda=3)"}
lambda <- 3
n_trials <- 10000  # fine-grained time steps
p_event <- lambda / n_trials  # tiny probability per step
n_reps <- 10000
set.seed(12345)

# "Manual": Poisson as the limit of a binomial with many tiny-probability trials
results_manual <- numeric(n_reps)
for (i in 1:n_reps) {
  results_manual[i] <- sum(rbinom(1, n_trials, p_event))
}

hist(results_manual, breaks = seq(-0.5, max(results_manual) + 1.5), freq = FALSE,
     main = "Binomial limit (n = 10000): rare events (lambda = 3)",
     xlab = "Number of events")
```

```{r poisson-shortcut, fig.cap="Poisson shortcut (dpois) overlaid on the binomial-limit simulation"}
# Shortcut: the Poisson formula, dpois(), evaluated at each possible count
k <- 0:max(results_manual)

hist(results_manual, breaks = seq(-0.5, max(results_manual) + 1.5), freq = FALSE,
     main = "Binomial limit (black bars) vs Poisson shortcut (red points)",
     xlab = "Number of events")
points(k, dpois(k, lambda), col = "red", pch = 19)
legend("topright", legend = c("Binomial-limit simulation", "dpois() shortcut"),
       col = c("black", "red"), pch = c(15, 19), bty = "n")
```

------------------------------------------------------------------------

# Example 3: Geometric — Trials until first success

**Process**: Flip the same biased coin (p=0.6) until first head. How many flips?

**Teaching moment**: The geometric distribution is the shortcut for counting trials until the first success.

```{r geometric-manual, fig.cap="Manual simulation: flips until first head (p=0.6)"}
p <- 0.6
n_reps <- 10000
max_trials <- 100
set.seed(12345)

# Manual: sample until first 1, count attempts
results_manual <- numeric(n_reps)
for (i in 1:n_reps) {
  flips <- sample(c(0, 1), max_trials, replace = TRUE, prob = c(1 - p, p))
  first_success <- which(flips == 1)[1]
  results_manual[i] <- ifelse(is.na(first_success), max_trials, first_success)
}

hist(results_manual, breaks = seq(0.5, max_trials + 0.5), freq = FALSE,
     main = "Manual: flips until first head (p = 0.6)",
     xlab = "Number of trials")
```

```{r geometric-shortcut, fig.cap="Geometric shortcut (dgeom) overlaid on manual simulation"}
# Shortcut: the geometric formula, dgeom().
# Watch the convention: dgeom() counts the number of FAILURES before the first
# success (0, 1, 2, ...), but our manual sim counts the TRIAL on which the first
# success occurs (1, 2, 3, ...). So P(first success on trial k) is dgeom(k - 1, p).
k <- 1:max_trials

hist(results_manual, breaks = seq(0.5, max_trials + 0.5), freq = FALSE,
     main = "Manual (black bars) vs Shortcut (red points)",
     xlab = "Number of trials")
points(k, dgeom(k - 1, p), col = "red", pch = 19)
legend("topright", legend = c("Manual simulation", "dgeom() shortcut"),
       col = c("black", "red"), pch = c(15, 19), bty = "n")
```

------------------------------------------------------------------------

# Example 4: Normal — a quantity built from many small effects

**Process**: A quantity is produced by *adding up* many small, independent random effects (here, 20 of them, each a tiny nudge up or down). What does the total look like?

**Teaching moment**: The normal distribution is the shortcut for a quantity built by **summing many small independent effects** — whatever the individual effects look like (the Central Limit Theorem).

*Note:* we keep this as a **process that produces one quantity** — like the count-producing processes above. Later, in the session on *sampling distributions*, we'll see that a sample mean is also such a sum, which is why means turn out normal too.

```{r clt-manual, fig.cap="Manual simulation: sum of 20 small independent effects"}
n_effects <- 20     # how many small effects add up to make the quantity
n_reps <- 10000
set.seed(12345)

# Manual: build a quantity by ADDING many small independent effects, then repeat
results_manual <- numeric(n_reps)
for (i in 1:n_reps) {
  effects <- runif(n_effects, -0.5, 0.5)  # many tiny random contributions
  results_manual[i] <- sum(effects)       # the process output is their sum
}

hist(results_manual, breaks = 30, freq = FALSE,
     main = "Manual: sum of 20 small independent effects",
     xlab = "Total effect")
```

```{r clt-shortcut, fig.cap="Normal shortcut (dnorm curve) overlaid on the sum-of-effects simulation"}
# Shortcut: the normal density curve, dnorm(), using the mean and sd OF THE SUM.
# For a sum of independent pieces, the means add and the variances add:
mu_total <- n_effects * 0         # each effect has mean 0
var_one  <- 1^2 / 12              # variance of one uniform(-0.5, 0.5) effect (width 1)
sd_total <- sqrt(n_effects * var_one)

hist(results_manual, breaks = 30, freq = FALSE,
     main = "Manual (black bars) vs Normal shortcut (red curve)",
     xlab = "Total effect")
curve(dnorm(x, mean = mu_total, sd = sd_total), col = "red", lwd = 2, add = TRUE)
legend("topright", legend = c("Manual simulation", "dnorm() shortcut"),
       col = c("black", "red"), lwd = c(6, 2), bty = "n")
```
