---
title: "Running R Code: A Hands-On Setup Session"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 4.5)
```

# What this session is for

This session is **not** about learning to write R code. It has one goal:

> Make sure you can **run** R code that I give you, **see its output**, and that
> R can **find the data files** it needs.

Later in the course I will hand you short scripts that illustrate the theory.
You will not need to write them — but you *will* need to open them, run them, and
read the output. This session sets that up.

**How to work through it.** Do the tasks below in order, on your own laptop.
After each task there is a **"You should see"** note. If what you see matches,
move on. If a task fails and the troubleshooting notes do not fix it, **bring your
laptop to class** and we will sort it out together.

---

# How this course works: one RStudio Project

Everything in this course lives inside a single **RStudio Project** — the folder I
give you, which contains an `.Rproj` file, a `data/` folder, and the scripts.

Why this matters, and why we insist on it:

- **Opening the project sets R's working directory to that folder.** That is what
  lets every script find its data with a short, relative name like
  `data/wages1833.csv` — the *same* path that works on my computer works on yours,
  with nothing to edit.
- You therefore **never** need `setwd()`, and **never** need a long,
  computer-specific path like `C:/Users/you/Documents/...`. If you find yourself
  writing one, stop — it means you have not opened the project.

**The one rule for this whole course:**

> **Always start by opening the class project** (double-click its `.Rproj` file),
> and keep your scripts inside the project folder. That is all it takes for the
> code I give you to just work.

---

# Task 0 — Open the class project

1. Find the folder I gave you for this course and, inside it, the file ending in
   **`.Rproj`**. **Double-click it** — RStudio opens with the project loaded.
2. Check the **top-right corner** of RStudio: it should show the **project's name**
   (not "Project: (None)"). In the **Files** pane (bottom-right) you should see the
   `data` folder.

Now confirm R is actually pointed at the project by running this one line — it
should print `TRUE`:

```{r task0-check}
file.exists("data/wages1833.csv")   # TRUE means R can see the project's data
```

**You should see:** `[1] TRUE`. If it prints `FALSE`, R is *not* in the project:
close RStudio and re-open it by **double-clicking the `.Rproj` file** (do not
launch R or RStudio on its own, and do not open the script by itself).

> **This is the foundation for everything below.** If Task 0 does not give you
> `TRUE`, stop and fix it first — none of the later tasks can find their data
> otherwise.

---

# Task 1 — Run a line of code

1. Make a new script: **File ▸ New File ▸ R Script**. Save it (**File ▸ Save**)
   as `myfirst.R` — **inside the project folder** (that is the location RStudio
   offers by default once the project is open).
2. Type or paste these lines into the script:

```{r task1, eval=FALSE}
2 + 2
x <- c(10, 20, 30, 40)
mean(x)
plot(1:10)
```

3. There are three ways to run code — try each:
   - **One line:** put the cursor on a line and press **Ctrl+Enter** (Mac: **Cmd+Enter**).
   - **Several lines:** select them with the mouse, then **Ctrl+Enter**.
   - **The whole file at once:** click the **Source** button (top of the editor).

**You should see:** `4` and `25` printed in the **Console** (bottom-left), and a
simple plot of ten points in the **Plots** pane (bottom-right).

> "Run" runs the current line or selection. "Source" runs the entire file from
> top to bottom — that is what you will do with the scripts I give you.

---

# Task 2 — Install and load the tidyverse

We will use the **tidyverse** — a collection of packages — to read data files
(and, later in the course, to wrangle them). You **install** a package once, then
**load** it at the top of each script that uses it.

First, install it. Run this line **once**; it is a large download, so be patient
(it can take a few minutes the first time):

```{r task2-install, eval=FALSE}
install.packages("tidyverse")   # run ONCE; big one-time download
```

Now load it. You will do this — or the scripts I give you will do this — every
time you start:

```{r task2-library, message=TRUE, warning=FALSE}
library(tidyverse)   # load the packages (put this at the top of a script)
```

**You should see:** a startup message that lists the attached packages
(`readr`, `dplyr`, `ggplot2`, …) and a few "Conflicts". That message is
**normal** — it is not an error.

> **Troubleshooting.** If you instead get
> `Error in library(tidyverse) : there is no package called 'tidyverse'`,
> you have not installed it yet. Run the `install.packages("tidyverse")` line
> above once, wait for it to finish, then try `library(tidyverse)` again.

---

# Task 3 — Read in a data file

This is the step that most often trips people up. Paste this line into your
script and run it:

```{r task3-read, message=TRUE}
wages <- read_csv("data/wages1833.csv")
```

**You should see:** a short message like `Rows: 51 Columns: 6` with the column
types — and in the **Environment** pane (top-right), `wages` now appears as
*51 obs. of 6 variables*. `read_csv()` gives you a **tibble**: a tidy table that
prints neatly.

### If instead you got an error

If you see a red message like:

```
Error: 'data/wages1833.csv' does not exist in current working directory:
'/Users/you/…'.
```

...then R is looking in the wrong folder (and helpfully prints which one) —
almost always because the project is not open. Fixes, in order:

1. Redo **Task 0** and confirm `file.exists("data/wages1833.csv")` prints `TRUE`.
   Nine times out of ten, this is the whole problem.
2. Check the **top-right** of RStudio shows the **project's name**, not
   "Project: (None)". If it says None, you opened R without the project.
3. Confirm a `data` folder really sits inside the project folder and contains
   `wages1833.csv`.

For reference, this is the wrong-path mistake — the file name without the `data/`
folder in front will fail:

```{r task3-wrong, eval=FALSE}
wages <- read_csv("wages1833.csv")   # WRONG: the data/ folder is missing
```

---

# Task 4 — Confirm the data loaded

Run these to look at what you just read in:

```{r task4-glimpse}
glimpse(wages)   # a compact overview: each column, its type, and first values
```

```{r task4-print}
wages            # a tibble prints its first rows neatly on its own
```

**You should see:** `glimpse` reporting `Rows: 51`, `Columns: 6`, with the age
and wage columns listed — and a tidy printout of the table. (The first column is
auto-named `...1`; it is just a row counter that was stored in the file.)

> `glimpse()` is how you check that *any* script loaded its data correctly.

---

# Task 5 — Run a complete demo, start to finish

This is a dress rehearsal for the theory scripts later in the course. Paste this
whole block into your script, select all of it, and run it (or click **Source**).
You do **not** need to understand the code — just confirm it runs.

```{r task5-demo, message=FALSE}
library(tidyverse)   # a script loads the packages it needs, up front

# Read the data (1833 Lancashire cotton-workers' wages by age)
wages <- read_csv("data/wages1833.csv")

# Print a couple of numbers to the Console
cat("Number of age groups:", nrow(wages), "\n")
cat("Average male wage:", round(mean(wages$mwage), 1), "\n")

# Draw one figure: how male wage changes with age
plot(wages$age, wages$mwage,
     type = "b", pch = 19, col = "steelblue",
     main = "Male wages by age (Lancashire, 1833)",
     xlab = "Age (years)", ylab = "Wage")
```

**You should see:** two lines printed in the Console, and a figure of wages rising
and then leveling off with age. If you got both, you can run a script end to end.

---

# Task 6 — Different scripts read different files

Not every data file uses commas, and different scripts read different files.
`readr` has a reader matched to each layout — `read_csv()` for commas,
`read_table()` for values lined up with spaces, `read_tsv()` for tabs. When I give
you a script, the right one is already chosen. Run each block and confirm a plot
appears.

**A larger data set (body measurements), commas — `read_csv`:**

```{r task6-body, message=FALSE}
body <- read_csv("data/body.csv")

plot(body$Height, body$Weight,
     pch = 19, col = "darkorange",
     main = "Weight vs height (body measurements)",
     xlab = "Height (cm)", ylab = "Weight (kg)")
```

**A file lined up with spaces (`cars.txt`) — `read_table`:**

```{r task6-cars, message=TRUE}
cars <- read_table("data/cars.txt")   # values separated by spaces, not commas

plot(cars$weight, cars$mpg_city,
     pch = 19, col = "seagreen",
     main = "Fuel use vs weight (cars)",
     xlab = "Weight", ylab = "City miles per gallon")
```

**You should see:** a scatter of weight against height (roughly increasing), and
a scatter showing heavier cars getting fewer miles per gallon. (`read_table` may
print a column-specification message — that is normal.)

---

# Task 7 — One-block self-check

Once everything above works, here is a single block that confirms your whole
setup at once. Paste it into a fresh script and **Source** it. If it runs to the
end and prints a small table, you are fully set up.

```{r task7-selfcheck, message=FALSE}
cat("R version:", R.version.string, "\n")

library(tidyverse)                          # do the packages load?
wages <- read_csv("data/wages1833.csv")     # does the data read?
cat("Rows read:", nrow(wages), "\n")        # should say 51
head(wages)                                 # does a small table print?
```

**You should see:** your R version, `Rows read: 51`, and a neat little table.
If all three appear with no red errors, you can stop here — you're ready.

---

# You're ready if...

Tick these off. You can, **without help**:

- [ ] Open the class project (`.Rproj`) and get `TRUE` from `file.exists("data/wages1833.csv")`
- [ ] Run a single line of code and see the answer in the Console
- [ ] Run a whole block at once (Source) and see its output
- [ ] Load the tidyverse with `library(tidyverse)` (no "no package" error)
- [ ] Read in a data file with **no** "does not exist" error
- [ ] See a plot appear in the Plots pane

If any box is unchecked, **come to class with your laptop** and we'll fix it
together — one issue at a time.
