---
title: "Sampling Distributions: How a Statistic Varies from Sample to Sample"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5)
```

# Overview

This lab builds directly on the distributions lab — **one level up**.

- There, we simulated a process and made a histogram of the **data**.
- Here, we simulate a process, compute a **summary statistic** (a mean, a median, …) from each sample, and make a histogram of the **statistic**.

That histogram-of-a-statistic is the **sampling distribution**. It is how we describe the *uncertainty* in a number computed from data.

The method is the same as before: **simulate many times, then overlay the closed-form shortcut** to show that the analytic result is the limit of the simulation.

---

# The setup: a known population

We invent a population so we know the truth and can check our answers. Deliberately, it is **not** bell-shaped — it is strongly right-skewed (a few very large values). Keep an eye on that: the population is skewed, but watch what the distribution of the sample *mean* turns out to look like.

```{r population}
set.seed(1)
population <- rexp(100000, rate = 1/50)  # 100,000 values; average around 50
pop_mean <- mean(population)
pop_sd   <- sd(population)

hist(population, breaks = 60, col = "grey80",
     main = "The population (strongly right-skewed)", xlab = "Value")
abline(v = pop_mean, col = "red", lwd = 3, lty = 2)
cat("Population mean:", round(pop_mean, 1), " sd:", round(pop_sd, 1), "\n")
```

---

# Three distributions you must keep apart

The single biggest source of confusion in this topic is mixing up three different distributions.

**1. The population** — shown above; the process that generates data.

**2. One sample** — draw `n = 25` values. This is *one dataset*. Its histogram is rough, and its mean is close to — but not exactly — the population mean.

```{r one-sample}
sample_size <- 25
one_sample <- sample(population, sample_size)

hist(one_sample, breaks = 12, col = "lightblue",
     main = "One sample (n = 25)", xlab = "Value")
abline(v = mean(one_sample), col = "blue", lwd = 3, lty = 2)
cat("This sample's mean:", round(mean(one_sample), 1),
    " (population mean:", round(pop_mean, 1), ")\n")
```

**3. The sampling distribution of the mean** — now repeat "draw a sample, compute its mean" many times, and histogram those means. *This* is the sampling distribution: the distribution of the statistic across repeated samples.

```{r sampling-dist}
n_reps <- 10000
sample_means <- numeric(n_reps)
for (i in 1:n_reps) {
  sample_means[i] <- mean(sample(population, sample_size))
}

hist(sample_means, breaks = 40, col = "grey80", freq = FALSE,
     main = "Sampling distribution of the mean (n = 25)",
     xlab = "Sample mean")
abline(v = pop_mean, col = "red", lwd = 3, lty = 2)
```

Notice two things. Even though the population is strongly skewed, the sampling distribution of the mean is nearly **bell-shaped**, centered on the true population mean — and it is far **narrower** than the population.

---

# The closed-form shortcut

Just like the distributions lab, the simulated histogram has a formula it converges to. The sampling distribution of the mean is approximately

$$\text{Normal}\left(\text{mean} = \mu,\ \ \text{sd} = \frac{\sigma}{\sqrt{n}}\right)$$

where $\mu$ and $\sigma$ are the population mean and sd. That spread — the sd of the sampling distribution — has a name: the **standard error (SE)**.

```{r closed-form}
se <- pop_sd / sqrt(sample_size)

hist(sample_means, breaks = 40, col = "grey80", freq = FALSE,
     main = "Simulation (bars) vs closed-form Normal (red curve)",
     xlab = "Sample mean")
curve(dnorm(x, mean = pop_mean, sd = se), col = "red", lwd = 2, add = TRUE)
legend("topright", legend = c("Simulated means", "Normal(mu, sigma/sqrt(n))"),
       col = c("grey40", "red"), lwd = c(6, 2), bty = "n")

cat("Standard error (formula sigma/sqrt(n)):", round(se, 2), "\n")
cat("SD of the simulated means            :", round(sd(sample_means), 2), "\n")
```

> **Callback to the distributions lab.** There, the normal was the shortcut for *a sum of many small effects*. A sample mean is just a sum divided by `n` — so sample means come out normal too. The new piece is the `/ sqrt(n)`: it says how the spread shrinks as the sample grows.

---

# What controls the spread? The sample size n

The sampling distribution is centered on the true parameter. Its **spread is the standard error, and the SE shrinks as `n` grows** — specifically, by a factor of `sqrt(n)`.

```{r n-dependence}
ns   <- c(5, 25, 100)
cols <- c("darkorange", "seagreen", "purple")

plot(NULL, xlim = c(0, 120), ylim = c(0, 0.09),
     main = "Sampling distribution of the mean for different n",
     xlab = "Sample mean", ylab = "Density")
abline(v = pop_mean, col = "red", lwd = 2, lty = 2)
for (j in seq_along(ns)) {
  means_j <- replicate(5000, mean(sample(population, ns[j])))
  lines(density(means_j), col = cols[j], lwd = 2)
}
legend("topright", legend = paste("n =", ns), col = cols, lwd = 2, bty = "n")
```

Larger `n` → narrower sampling distribution → a more precise estimate. Let's confirm the `sqrt(n)` law directly by simulating the SE at several sample sizes and overlaying the formula:

```{r se-law}
ns2    <- c(5, 10, 25, 50, 100, 200)
se_sim <- sapply(ns2, function(nn) sd(replicate(3000, mean(sample(population, nn)))))

plot(ns2, se_sim, pch = 19, col = "seagreen", cex = 1.3,
     main = "Standard error vs sample size",
     xlab = "Sample size n", ylab = "SE (spread of the sample mean)")
curve(pop_sd / sqrt(x), col = "red", lwd = 2, add = TRUE)
legend("topright", legend = c("Simulated SE", "Formula sigma/sqrt(n)"),
       col = c("seagreen", "red"), pch = c(19, NA), lwd = c(NA, 2), bty = "n")
```

---

# The mean is not special: every statistic has a sampling distribution

We used the mean, but the same idea applies to *any* summary you compute from a sample. Below are the sampling distributions of four different statistics, all from the same population at `n = 25`.

```{r other-stats}
n_reps <- 5000
stat_mean   <- replicate(n_reps, mean(sample(population, sample_size)))
stat_median <- replicate(n_reps, median(sample(population, sample_size)))
stat_sd     <- replicate(n_reps, sd(sample(population, sample_size)))
stat_prop   <- replicate(n_reps, mean(sample(population, sample_size) > 50))  # proportion above 50

par(mfrow = c(2, 2))
hist(stat_mean,   breaks = 30, col = "grey80", main = "Sampling dist: mean",           xlab = "")
hist(stat_median, breaks = 30, col = "grey80", main = "Sampling dist: median",         xlab = "")
hist(stat_sd,     breaks = 30, col = "grey80", main = "Sampling dist: sd",             xlab = "")
hist(stat_prop,   breaks = 30, col = "grey80", main = "Sampling dist: proportion > 50", xlab = "")
par(mfrow = c(1, 1))
```

Each statistic has its **own** sampling distribution with its own shape — the mean and median look roughly normal, the sd is right-skewed, the proportion is discrete. Some of these shapes have their own named closed forms — the **t**, **chi-squared**, and **F** distributions — which we will meet in later sessions.

---

# Why this matters: from an estimate to an inference

A single sample gives you a single number — an estimate. On its own, you cannot tell how much to trust it. The **sampling distribution is what turns that estimate into an inference**:

- its **center** tells you the estimate is, on average, on target;
- its **spread — the standard error —** tells you how much the estimate would wobble from one sample to the next.

This is the foundation for everything that follows:

- **Hypothesis testing** asks where an observed statistic falls in a sampling distribution built *assuming no effect* (a "null" sampling distribution).
- **Confidence intervals** turn the standard error into a range of plausible parameter values.

We build those in later sessions. For now, the key skill is the one you just watched: **to describe the uncertainty in any statistic, simulate its sampling distribution — or use the closed-form shortcut when one exists.**
